### 2020-06-02 writeup/intro.tex
sec 12

\begin{abstract}
When does multi-task learning provide positive (or negative) information transfer, compared to single-task learning?
In this work, we address this question by studying
% the performance of predicting a particular task given multiple tasks.
a multi-task learning architecture that uses a shared layer for all tasks and a separate output layer for each task.
First, for the setting of high-dimensional linear regression, we show that the type of transfer is determined by the bias-variance tradeoff of multi-task learning estimators.
We provide a sharp analysis of the tradeoff and extend the result to transfer learning.
We develop a technical tool on the inverse of the sum of two random matrices that naturally arises from the analysis.
Second, we identify three components including task similarity, data size and covariate shift to explain negative transfer in multi-task learning rigorously.
Finally, we show practical implications of our theory for detecting and mitigating negative transfer.
We propose an incremental training scheduler for improving the efficiency of multi-task training for predicting a particular task.
We validate our theory and our proposed method on image and text classification tasks.
\end{abstract}

@@

\textbf{Setup and Main Results.}
We consider the high-dimensional linear regression setting \cite{HMRT19,BLLT20} to explain the above phenomena.
We focus on predicting a particular task that only contains limited amount of labeled data.
%Following Hastie et al.  and Bartlett et al. \cite{},
We assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We focus on the hard parameter sharing architecture with a linear shared layer for all tasks and a separate prediction head for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_setup}.
Our main theoretical contribution is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

@@

%Concretely, we show a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which
Theorem \ref{lem_cov_shift_informal} extends a well-known result for the single-task setting when $X_1, \rho_1, a_1 = 0$ \cite{S07}.
The concentration error of our result, which is on the order of $O(p^{-1/2+\varepsilon})$, is nearly optimal.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
%We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.
Based on Theorem \ref{lem_cov_shift_informal}, we provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) many tasks with the same features (e.g. images).
We extend our result to the transfer learning
%in the setting of high-dimensional linear regression.
%by pooling source task representations into the shared body of the hard parameter sharing architecture, following
setting of Taskonomy by Zamir et al. \cite{ZSSGM18}.
We prove that the bias of the transfer learning estimator is given by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are described more precisely in Section \ref{sec_main}.

Our next contribution is to use our newly developed tool to explain negative transfer in multi-task learning precisely.
In Section \ref{sec_insight}, we explain the three phenomena in Figure \ref{fig_model_shift_phasetrans} for a simplified isotropic setting of two tasks.
It is crucial that the concentration error of Theorem \ref{lem_cov_shift_informal} is $O(p^{-1/2+\varepsilon})$ so that we can explain the transition phenomena in Figure \ref{fig_model_shift} and \ref{fig_size}.
The unexplained observations are caused by an error term that arises from the bias of $\hat{\beta}_t^{\MTL}$ -- we discuss these in Section \ref{sec_insight} more precisely.
Theorem \ref{lem_cov_shift_informal} allows use to compare MTL performance under different covariate shift, measured by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$.
When $\beta_1 = \beta_2$, we show that for a certain family of bounded $M$ matrices, having no covariate shift, i.e. all singular values of $M$ being equal, yields the optimal MTL performance.
%In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
%We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
%Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
%We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
Finally, we analyze the benefit of MTL for reducing the amount of labeled data needed to achieve comparable performance to STL, which is a key empirical finding of Taskonomy.
%We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
%		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

Our last contribution is to connect our theory to practical problems of interest.
(i) We provide a single-task based metric to predict positive or negative transfer in multi-task learning.
While it is not well understood when multi-task learning provides positive transfer, we show that the STL results can help indicate and understand MTL results on ChestX-ray14 \cite{chexnet17} and sentiment analysis datasets \cite{LZWDA18}.
(ii) We design an incremental training schedule to improve the efficiency of multi-task training for predicting a particular task.
We show that our training schedule reduces the computational cost by $55\%$ compared to baseline multi-task training on the sentiment analysis dataset, while keeping the accuracy the same.
(iii) We validate the three components of our theory in Section \ref{sec_insight} and measure the data efficiency of multi-task learning on the sentiment analysis dataset.

>>>>

\begin{abstract}
When does multi-task learning provide positive (or negative) information transfer, compared to single-task learning?
In this work, we address this question by studying
% the performance of predicting a particular task given multiple tasks.
a multi-task learning architecture that uses a shared layer for all tasks and a separate output layer for each task.
First, for the setting of high-dimensional linear regression, we show that the type of transfer is determined by the bias-variance tradeoff of multi-task learning estimators.
We provide an analysis of the tradeoff and extend the result to transfer learning.
We develop a technical tool on the inverse of the sum of two random matrices that naturally arises from the analysis.
Second, we identify three components including task similarity, data size and covariate shift to explain negative transfer in multi-task learning rigorously.
Finally, we show practical implications of our theory for detecting and mitigating negative transfer.
We propose an incremental training scheduler for improving the efficiency of multi-task training for predicting a particular task.
We validate our theory and our proposed method on image and text classification tasks.
\end{abstract}

@@

%\textbf{Setup and Main Results.}
We consider the high-dimensional linear regression setting \cite{HMRT19,BLLT20} to explain the above phenomena.
We focus on predicting a particular task that only contains limited amount of labeled data.
%Following Hastie et al.  and Bartlett et al. \cite{},
We assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We focus on the hard parameter sharing architecture with a linear shared layer for all tasks and a separate prediction head for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_setup}.
\textit{Our first contribution} is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

@@

%Concretely, we show a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which
Theorem \ref{lem_cov_shift_informal} extends a well-known result for the single-task setting when $X_1, \rho_1, a_1 = 0$ \cite{S07}.
The concentration error of our result, which is on the order of $O(p^{-1/2+\varepsilon})$, is nearly optimal.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
%We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.
Based on Theorem \ref{lem_cov_shift_informal}, we provide an analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) many tasks with the same features (e.g. images).
We extend our result to the transfer learning
%in the setting of high-dimensional linear regression.
%by pooling source task representations into the shared body of the hard parameter sharing architecture, following
setting of Taskonomy by Zamir et al. \cite{ZSSGM18}.
We prove that the bias of the transfer learning estimator is given by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
These results are described more precisely in Section \ref{sec_main}.

\textit{Our next contribution} is to use our newly developed tool to explain negative transfer in multi-task learning precisely.
In Section \ref{sec_insight}, we explain the three phenomena in Figure \ref{fig_model_shift_phasetrans} for a simplified isotropic setting of two tasks.
It is crucial that the concentration error of Theorem \ref{lem_cov_shift_informal} is $O(p^{-1/2+\varepsilon})$ so that we can explain the transition phenomena in Figure \ref{fig_model_shift} and \ref{fig_size}.
The unexplained observations are caused by an error term that arises from the bias of $\hat{\beta}_t^{\MTL}$ -- we discuss these in Section \ref{sec_insight} more precisely.
Theorem \ref{lem_cov_shift_informal} allows us to compare MTL performance under different covariate shift, measured by $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$.
When $\beta_1 = \beta_2$, we show that for a certain family of bounded $M$ matrices, having no covariate shift, i.e. all singular values of $M$ being equal, yields the optimal MTL performance.
%In Section \ref{sec_insight}, we consider three components including task similarity, data size and covariate shift for a simplified isotropic setting of two tasks.
%We measure task similarity by how small is the distance between $\beta_1$ and $\beta_2$.
%Using our tool, we explain a transition from positive to negative transfer as task similarity decreases.
%		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
%We show that increasing the data size of the source task does not always improve performance for the target task in multi-task learning.
Finally, we analyze the benefit of MTL for reducing the amount of labeled data needed to achieve comparable performance to STL, which is a key empirical finding of Taskonomy.
%We show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
%		We show that as $n_1 / n_2$ becomes large, having no covariate shift between the source and target tasks yields the optimal performance for the target task.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.

\textit{Our last contribution} is to connect our theory to practical problems of interest.
(i) We provide a single-task based metric to predict positive or negative transfer in multi-task learning.
While it is not well understood when multi-task learning provides positive transfer, we show that the STL results can help indicate and understand MTL results on ChestX-ray14 \cite{chexnet17} and sentiment analysis datasets \cite{LZWDA18}.
(ii) We design an incremental training schedule to improve the efficiency of multi-task training for predicting a particular task.
We show that our training schedule reduces the computational cost by $55\%$ compared to baseline multi-task training on the sentiment analysis dataset, while keeping the accuracy the same.
(iii) We validate the three components of our theory in Section \ref{sec_insight} and measure the data efficiency of multi-task learning on the sentiment analysis dataset.
